Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-02-22 paperDistance/paper.tex
discussion content

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ data without any use of physical stellar
models.
Although we have not performed principled Bayesian inference in this
project, we have performed a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the model we use (Extreme Deconvolution or XD) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the color--magnitude diagram (CMD)---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the CMD we have built is representative for
the \tgas\ Catalog.
Inasmuch as that is true, we expect the distance estimates to be
weakly biased.

>>>>

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
We obtained these by building a data-driven model of the
color--magnitude diagram (CMD) of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian interences.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown thatm, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters are much more clustered than likelihood-based distance
estimates.

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's distance by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the model we use (Extreme Deconvolution or XD) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the color--magnitude diagram (CMD)---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the CMD we have built is representative for
the \tgas\ Catalog.
Inasmuch as that is true, we expect any (sensible) distance estimates
we generate from our distance posterior pdfs to be weakly biased.
